Exploitation Techniques and Mitigations

### 2016-01-26 docs/paper.tex
Paper: Draft CFI

We happily receive a shell upon running the exploit. Execution will be handed
back to the original binary after we close the shell. Since we messed up the
control flow with our exploit the program segfaults shortly after.

@@

DEP can be fooled by ROP and ASLR is rendered useless with a simple info leak.
Something else is required at this point. Thinking back, the original problem
emerged from manipulating the return address located on the stack. Two
(additional) counter mechanisms were introduced going by the names StackGuard
and StackShield. We will take only a look at StackGuard and one relatively
common scenario to break it, but there is a comprehensive article
\cite{phrack_stack_guard} on \texttt{phrack}\footnote{\url{http://phrack.org/}}
describing and breaking both mechanisms.

@@

\section{Control Flow Integrity (CFI)}

>>>>

We happily receive a shell upon running the exploit. Execution will be handed
back to the original binary after we close the shell. Since we messed up the
control-flow with our exploit the program segfaults shortly after.

@@

DEP can be fooled by ROP and ASLR is rendered useless with a simple info leak.
Something else is required at this point. Thinking back, the original problem
emerged from manipulating the return address located on the stack. Two
(additional) counter mechanisms were introduced going by the names of
StackGuard and StackShield. We will take only a look at StackGuard and one
relatively common scenario to break it, but there is a comprehensive article
\cite{phrack_stack_guard} on \texttt{phrack}\footnote{\url{http://phrack.org/}}
describing and breaking both mechanisms.

@@

\section{Control-Flow Integrity (CFI)}

In this section we are going to have a short glimpse at Control-Flow Integrity,
but before that we need to talk about the Control-Flow Graph (CFG).

\subsection{Control-Flow Graph (CFG)}

Again the name already tells you what this is about, a direct acyclic graph
(DAG) that reflects the control-flow of your program. There are different
definitions regarding what is actually placed in the nodes of the graph and
what are the anchor points of the program used to connect them. In our case we
will create a node out of each function body and connect them at function
calls.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{gfx/unraid_part_graph.png}
    \caption{Truncated control-flow graph example}
    \label{fig:cfg}
\end{figure}

\Cref{fig:cfg} shows part of such a graph in
IDA\footnote{\url{https://www.hex-rays.com/products/ida/}} analyzing the
\texttt{emhttp} binary used by
unRAID\footnote{\url{http://lime-technology.com/what-is-unraid/}}. There are
much nicer graphs available to explain the concept, but we wanted to inject
some real world data. From the graph you can already tell that the function
(subroutine) labeled \texttt{licensing} will call \texttt{malloc},
\texttt{memset}, \texttt{USB\_STUFF}, \dots. Thats actually all we care about
for now.

\subsection{Back to CFI}

CFI is a big topic and, like other topics already mentioned, goes beyond the
scope of this writeup. The first pointer we will hand you aims at corresponding
section\footnote{\url{http://clang.llvm.org/docs/ControlFlowIntegrity.html}} of
the Clang documentation, but we encourage you to checkout the related research
paper \cite{ms_cfi} for more information. A more accessible and recent way to
this topic may be the
talk\footnote{\url{https://www.youtube.com/watch?v=FA0VK7s5tSQ}} \textit{New
memory corruption attacks: why can't we have nice things?} given by gannimo
(Mathias Payer).

The CFG has already been established, now let's see how it can be used to
counter the buffer overflow return address dilemma. At compile-time the graph
is available and can be used to create additional constraints which the program
must obey during runtime. This is similar to the StackGuard mechanism where we
attach code to the end of a function which checks if the canary is still
intact. But now we don't check for a canary but for the validity of the return
address. From the CFG we can build a set of possible return targets for each
function. Looking back at the example shown in \cref{fig:cfg} we can determine
that the function \texttt{szDecrypt} returns either to \texttt{USB\_STUFF},
\texttt{licensing} or one other subroutine excluded by the illustration. The
corresponding addresses are put into this set which is then stored in the
binary. Before the function \texttt{szDecrypt} returns the return address on
the stack is compared to the entries listed in the corresponding set. If no
match can be found, the program terminates.

With this mechanism setup, one can easily see that it is no longer possible
chain \emph{arbitrary} gadgets together to pull off ROP. But we still control
the return address and can jump to different locations as long as we stick to
the CFG. Each transition from one node to another has to be valid, while the
overall path taking by our ROP chain may do things never intended by the
program's author. This concept is known as control-flow bending.~
\cite{cf_bending}

\subsection{Stack Integrity}

Using a changed return address is what ultimately enables control-flow bending.
Stack integrity ensures that the same return address is used upon executing the
\texttt{ret} instruction as was pushed upon function call. This can be achieved
by using a \textit{shadow-stack} similar to the StackShield mechanism but we
suggest reading about \textit{code-pointer integrity} \cite{cpi} regarding this
topic.

\subsection{There is an interpreter in your C}

We conclude this section by mentioning the availability of an interpreter
(probably) available in your standard library. As presented by gannimo,
\texttt{printf} is far more capable than just printing arguments. It is also
possible to read and \emph{write} to memory locations. But one can go even
further and craft a format string mimicking each of the eight operators of
Brainfuck\footnote{\url{https://en.wikipedia.org/wiki/Brainfuck}}. Because
Brainfuck is Turing complete we can deduce that \texttt{printf} is a Turing
complete interpreter. Note that this requires \texttt{printf} to be called in a
loop and the format strings may depend heavily on your libraries
implementation. (Also not all implementation are Turing complete).

A compiler accepting Brainfuck and spitting out the corresponding format
strings, including examples can be found on HexHive's
GitHub\footnote{\url{https://github.com/HexHive/printbf}}.
